---
title: "BST270_final"
author: "Arpan Sarkar"
date: "1/20/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE,
                      tidy.opts=list(width.cutoff=80))
```

# Overview

In this project, I aim to reproduce the COVID-19 visualizations and tables published by the New York Times. Specifically, I attempt to reproduce the following for January 17th, 2021:

1. New cases as a function of time with a rolling average plot
2. Table of cases, hospitalizations and deaths
3. Table of cases by state

# Load relevant packages

```{r}
library(ggplot2)
library(tidyverse)
library(stringr)
library(zoo)
library(lubridate)
library(kableExtra)
```

# Data

Data for cases and deaths can be downloaded from this NYT GitHub repository (see us-counties.csv and us.csv). Data for hospitalizations can be downloaded from The COVID Tracking Project (see all-states-history.csv).

```{r}
states <- read.csv("Data/all-states-history.csv")
us <- read.csv("Data/us.csv")
us_s <- read.csv("Data/us-states.csv")
```

# Figure 1: New cases as a function of time with a rolling average plot

This figure features a graph of new Covid-19 cases per day with a rolling 7-day average plot of new Covid-19 cases.

As presented in the us.csv data, cases are cumulative, so it is necessary to create a new column with values for new cases for each recorded day.

```{r}
us$new_cases <- us$cases - lag(us$cases)
us$new_cases_7dayavg = rollmean(us$new_cases, k = 7, fill = NA,align="right")
```

In order to recreate the x-axis of the NYT figure more faithfully, I format the date column by month.

```{r}
us$date <- ymd(us$date)
us$month <- format(us$date,"%B")
```

The plot itself is created as follows:

```{r}
plot <- ggplot(us, aes(date, new_cases)) + geom_col(aes(date, new_cases),color="lightpink1",alpha=0.8) +
  geom_path(aes(date,new_cases_7dayavg),color="darkred") +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000), labels = c("0", "100,000", "200,000", "300,000"))+
  xlab("Date")+
  ylab("Cases")
print(plot)
```

**Notes on Reproducibility:** This honestly seemed pretty clear from the figure provided in the problem specifications. It was clear that we were using new cases and the 7-day averages were fairly clear.


# Figure 2: Table of cases, hospitalizations and deaths

This figure features a table with total reported Covid-19 cases, total reported Covid-19 deaths, new recorded Covid-19 cases on Jan. 17th, new recorded Covid-19 deaths on Jan. 17th, Covid-19 hospitalizations on Jan. 17th, 14-day change for Covid-19 cases, 14-day change for Covid-19 deaths, and 14-day change for Covid-19 hospitalizations. 

The total reported Covid-19 cases and Covid-19 deaths as of Jan. 17th are present in the original (un-modified) us.csv data set, and the number of current hospitalizations as of Jan. 17th is present in the original all-states-history.csv dataset. The new recorded Covid-19 cases on Jan. 17th was calculated in the process of generating Figure 1. The new recorded Covid-19 deaths on Jan. 17th can be calculated in a similar process:

```{r}
us$new_deaths  <- us$deaths - lag(us$deaths)
us$new_deaths_7dayavg = rollmean(us$new_deaths, k = 7, fill = NA, align="right")
```

In order to determine the 14-day changes (in percent), the values new cases, deaths, and current hospitalizations must be recorded for Jan. 3rd and Jan 17th. (14 days prior).

```{r}
Jan3_cd <- us[us$date == "2021-01-03",]
Jan17_cd <- us[us$date == "2021-01-17",]
```

The current hospitalizations should be grouped by date since otherwise the values provided are for each individual state/territory on any given day.

```{r}
total_hospitalizedCurrently = states %>%
  group_by(date) %>%
  summarize(totalnew_hosps = sum(hospitalizedCurrently, na.rm=T)) 
total_hospitalizedCurrently$date <- rev(unique(states$date))
```

Calculations of 14-day percent changes of new cases, deaths, and current hospitalizations. It is important to note that this uses 7-day averages for the calculations.


```{r}
total_hospitalizedCurrently$totalnew_hosps_7dayavg <- rollmean(total_hospitalizedCurrently$totalnew_hosps, k = 7, fill = NA, align = "right")
Jan3_hosp <- total_hospitalizedCurrently[total_hospitalizedCurrently$date == "2021-01-03",]
Jan17_hosp <- total_hospitalizedCurrently[total_hospitalizedCurrently$date == "2021-01-17",]
```


```{r}
cases_change <- (Jan17_cd$new_cases_7dayavg - Jan3_cd$new_cases_7dayavg)/Jan3_cd$new_cases_7dayavg * 100
deaths_change <- (Jan17_cd$new_deaths_7dayavg - Jan3_cd$new_deaths_7dayavg)/Jan3_cd$new_deaths_7dayavg * 100
hosp_change <- (Jan17_hosp$totalnew_hosps_7dayavg - Jan3_hosp$totalnew_hosps_7dayavg)/Jan3_hosp$totalnew_hosps_7dayavg * 100
```

Completing the table:

```{r}
Cases <- round(c(Jan17_cd$cases,Jan17_cd$new_cases, cases_change),digits=0)
Deaths <- round(c(Jan17_cd$deaths, Jan17_cd$new_deaths, deaths_change),digits=0)
Hospitalized <- c("",round(Jan17_hosp$totalnew_hosps,digits=0),round(hosp_change,digits=0))
table <- as.matrix(rbind(Cases, Deaths, Hospitalized))
colnames(table) <- c("TOTAL REPORTED","ON JAN 17", "14-DAY CHANGE")
kable(table,digits=2)
```

**Notes on Reproducibility:** This also seemed pretty clear from the figure provided in the problem specifications. It was not immediately obvious to me that there was a gray caption under the table that details that the 14-day changes uses 7-day averages.

# Figure 3: Table of cases by state

This figure features a table with the total Covid-19 cases and daily average in the last 7 days of Covid-19 cases grouped by state.

It is necessary to compute the new cases for each state (we already did this for us.csv, but it did not have the state labels).

```{r}
us_s <- us_s %>%
  group_by(state) %>%
  arrange(date) %>%
  mutate(new_cases = cases - lag(cases))
```

Again, the rolling 7-day averages should be computed.

```{r}
us_s <- us_s %>% 
  group_by(state) %>% 
  mutate(new_cases_7dayavg = rollmean(new_cases, k = 7, fill = NA, align = "right"))
```

Only select the data for Jan. 17th.

```{r}
Jan17_3 <- us_s[us_s$date=="2021-01-17",c("state","cases","new_cases_7dayavg")]
names(Jan17_3) <- c("State","Total Cases", "Daily Avg in Last 7 Days")
Jan17_3$`Daily Avg in Last 7 Days` <- round(Jan17_3$`Daily Avg in Last 7 Days`)
```

Completing the table:

```{r}
options(scipen=999)
kable(Jan17_3)
```

**Notes on Reproducibility:** This third figure also seemed pretty clear from the figure provided in the problem specifications. This was perhaps the easiest of the figures to reproduce.